# Environmental predictability & trends

We applied harmonic regression to datasets containing the following variables:

1. `NDVI` *Normalized Difference Vegetation Index* from the [GIMMS NDVI from AVHRR Sensors 3rd Generation](https://developers.google.com/earth-engine/datasets/catalog/NASA_GIMMS_3GV0) dataset
1. `AET` *Actual Evapotranspiration* from the [TerraClimate](https://developers.google.com/earth-engine/datasets/catalog/IDAHO_EPSCOR_TERRACLIMATE) dataset
1. `PR` *Precipitation accumulation* from the [TerraClimate](https://developers.google.com/earth-engine/datasets/catalog/IDAHO_EPSCOR_TERRACLIMATE) dataset
1. `TMMX` *Maximum temperature* from the [TerraClimate](https://developers.google.com/earth-engine/datasets/catalog/IDAHO_EPSCOR_TERRACLIMATE) dataset
1. `TMMN` *Minimum temperature* from the [TerraClimate](https://developers.google.com/earth-engine/datasets/catalog/IDAHO_EPSCOR_TERRACLIMATE) dataset

All datasets contain monthly aggregates.

```{r setup_envpred_preprocessing_environment, include=FALSE}
library(stars)
library(raster)
library(tidyverse)
```

## Combine environmental datasets, trend datasets and ecoregions

We need to combine the predictors (environmental factors) and response (assemblage trends) to a single dataset before modelling. We will also limit the data spatially by the outline of the African continent (including Madagascar).

```{r combine_environmental_and_trend_datasets, results='hold'}
africa_outline <- st_read("data/raw/Africa.gpkg")
lut <- readRDS("data/processed/ecoregions_lut.RDS")

load_raster <- function(rasterpath, bandnames) {
  r <- stack(rasterpath)
  if (!is.null(bandnames)) {
    band_ids <- which(names(r) %in% bandnames)
    lapply(band_ids, function(x) raster(rasterpath, band = x))
  } else {
    unstack(r)
  }
}

ndvi <- load_raster("data/raw/gee/ndvi_1.tif", c("ndvi_original", "residuals", "slope"))
pr <- load_raster("data/raw/gee/pr_1.tif", c("pr_original", "residuals", "slope"))
pet <- load_raster("data/raw/gee/pet_1.tif", c("pet_original", "residuals", "slope"))
aet <- load_raster("data/raw/gee/aet_1.tif", c("aet_original", "residuals", "slope"))
tmmn <- load_raster("data/raw/gee/tmmn_1.tif", c("tmmn_original", "residuals", "slope"))
tmmx <- load_raster("data/raw/gee/tmmx_1.tif", c("tmmx_original", "residuals", "slope"))
def <- load_raster("data/raw/gee/def_1.tif", c("def_original", "residuals", "slope"))
realms <- load_raster("data/processed/realms.tif", NULL)
biomes <- load_raster("data/processed/biomes.tif", NULL)
ecoregions <- load_raster("data/processed/ecoregions.tif", NULL)

trend_long_assemblage <- raster::raster("data/processed/trend_long_assemblage.tif")
trend_short_assemblage <- raster::raster("data/processed/trend_short_assemblage.tif")
trend_long_counts <- raster::raster("data/processed/trend_long_counts.tif")
trend_short_counts <- raster::raster("data/processed/trend_short_counts.tif")

bandnames <- c("ndvi", "ndvi_resid", "ndvi_trend", "pr", "pr_resid", "pr_trend", "pet", "pet_resid", "pet_trend", 
               "aet", "aet_resid","aet_trend", "tmmn", "tmmn_resid", "tmmn_trend", "tmmx", "tmmx_resid", "tmmx_trend", 
               "def", "def_resid", "def_trend", "trend_long", "trend_short", "count_long", "count_short",
               lut$realms, lut$biomes, lut$ecoregions)

data <- suppressWarnings(brick(c(ndvi, pr, pet, aet, tmmn, tmmx, def, trend_long_assemblage, trend_short_assemblage, 
                                 trend_long_counts, trend_short_counts, realms, biomes, ecoregions)))
names(data) <- bandnames

data_stars <- st_as_stars(data)[africa_outline] %>%
  st_set_dimensions(3, values = bandnames) %>%
  st_set_dimensions(names = c("x", "y", "var"))
```

To confirm that worked, let's plot the output.

```{r plot_stars_data, out.width='100%'}
plot(data_stars)
```

## Save modeling dataset

We will now save the `stars` brick and a dataframe for modeling.

```{r save_modeling_datasets}
saveRDS(data_stars, file = "data/processed/data_stars.RDS")

as.data.frame(data_stars) %>%
  `colnames<-`(c("x", "y", "variable", "value")) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  drop_na(!any_of(c(lut$ecoregions, lut$realms, lut$biomes))) %>%
  identity() -> data

saveRDS(data, file = "data/processed/data.RDS")
```

